---
title: "Normalization and QC"
author: "Nicolas Veland"
format: html
editor: visual
---

# Normalization and QC (PCA)

This script starts with the output data generated from **01_processing.qmd** script.

This script performs the following steps:

1.  Use DESeq2 to normalize raw counts data using vst.
2.  Correct for batch effect.
3.  QC data inspection by PCA
4.  Save normalized count data matrix
5.  Add gene information
6.  Export normalized count wide dataframe.

```{r}
# Load Libraries
#| echo: false
#| message: false
library(DESeq2)
library(limma)
library(org.Mm.eg.db)
library(dplyr)
```

```{r}
######### II. Normalize using DESeq2 ----------------------------------------------

# 1. Create a DESeqDataSet object:
# design= ~ batch + condition --> The design indicates how to model the samples, here, that we want to measure the effect of the condition, controlling for batch differences. The two factor variables batch and condition should be columns of coldata.
# Note: In order to benefit from the default settings of the package, you should put the variable of interest at the end of the formula and make sure the control level is the first level.
dds = DESeqDataSetFromMatrix(
    countData = Gut_Raw_Counts,
    colData = Gut_colData,
    design= ~ Genotype)
```

```{r}
#  Plot a PCA to check batch effect
vsd_with_batch <- vst(dds)
plotPCA(vsd_with_batch, "Batch")
PCAwithBatch <- plotPCA(vsd_with_batch, "Genotype")

WTvsKO_colors = c("grey40", "red4") #c("salmon","grey40")

PCAwithBatch + 
  geom_text(aes(label = colnames(vsd_with_batch)), show.legend = FALSE, size = 5, vjust = "inward", hjust = "inward", position = position_nudge(y = 1.3)) +
  theme(text=element_text(size=15), legend.text=element_text(size=8)) +
  theme_bw(base_size = 15) +
  ggtitle("PCA with batch effect") +
  #theme(legend.position = c(0.1, 0.1)) +
  theme(legend.background=element_blank()) +
  scale_colour_manual(values= WTvsKO_colors,
                     name="Genotype",
                     breaks=c("WT", "KO"),
                     labels=c("WT", "KO")
                     )
```

```{r}
# 2. Remove Batch effect using limma package
vsd_without_batch = vst(dds)
assay(vsd_without_batch) <- limma::removeBatchEffect(assay(vsd_with_batch), vsd_with_batch$Batch)

```

```{r}
# 3. Plot a PCA for QC after removing batch effect 
plotPCA(vsd_without_batch, "Batch")

PCAwithoutBatch <- plotPCA(vsd_without_batch, "Genotype")
PCAwithoutBatch + 
  geom_text(aes(label = colnames(vsd_without_batch)), show.legend = FALSE, size = 5, vjust = "inward", hjust = "inward", position = position_nudge(x = 0, y = 2)) +
        theme(text=element_text(size=15), legend.text=element_text(size=8)) +
        theme_bw(base_size = 15) + coord_fixed(ratio = 0.75) +
        ggtitle("PCA without batch effect") +
        #theme(legend.position = c(0.1, 0.1)) +
        theme(legend.background=element_blank()) +
        scale_colour_manual(values= WTvsKO_colors,
                     name="Genotype",
                     breaks=c("WT", "KO"),
                     labels=c("WT", "KO")
                     )
```

```{r}
# Save PCA plot after removing batch effect
pdf(paste0(output_dir,"TEST_PCA_removedBatch.pdf"), 
    width = 7, height = 4, pointsize = 10)
print(PCAwithoutBatch + 
        geom_text(aes(label = colnames(vsd_without_batch)), show.legend = FALSE, size = 5,
                  vjust = "inward", hjust = "inward", 
                  position = position_nudge(x = 0, y = 1.2)) +
        theme(text=element_text(size=15), legend.text=element_text(size=8)) +
        theme_bw(base_size = 15) + coord_fixed(ratio = 0.75) +
        ggtitle("TEST PCA without batch effect") +
        #theme(legend.position = c(0.1, 0.1)) +
        theme(legend.background=element_blank()) +
        scale_colour_manual(values= WTvsKO_colors,
                     name="Genotype",
                     breaks=c("WT", "KO"),
                     labels=c("WT", "KO")
                     ))
dev.off()
```

```{r}
# 4. Write batch corrected normalized vst matrix for downstream plotting (e.g. for Boxplots or Heatmaps using vst counts):
Gut_normalized_withoutBatch <- assay(vsd_without_batch)



# 5. Modify counts matrix to add gene names

# a) Create geneid dataframe to assign gene names to Ensemble_id
Gut_normalized_withoutBatch = as.data.frame(Gut_normalized_withoutBatch)
Gut_normalized_withoutBatch <- tibble::rownames_to_column(Gut_normalized_withoutBatch, "Ensembl_id")

geneid = data.frame(
  Ensembl = Gut_normalized_withoutBatch$Ensembl_id,
  symbol = NA
)

# b) Get gene symbols from org.Mm.eg.db
geneid$symbol = mapIds(
  org.Mm.eg.db,
  keys = Gut_normalized_withoutBatch$Ensembl_id,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

# c) Check all Ensembl_ids in geneid dataframe matches the gene_ids in counts dataframe
all(Gut_normalized_withoutBatch$Ensembl_id == geneid$Ensembl
    )

# d) Insert gene name comlumn in counts dataframe from geneid dataframe symbol column
Gut_normalized_withoutBatch$gene = geneid$symbol

# From this point, counts dataframe has gene name to facilitate searching genes #<-------
# Check that TCR genes are annotated correctly and double check with Ensembl ID (e.g. ENSMUSG00000076746 = Trgv6). If not present, may need to be added manually.
Gut_normalized_withoutBatch$gene[21741] = "Trgv6"


```

### From this point, counts dataframe has gene name to facilitate searching genes

```{r}
# 6. Exporting Normalized Counts as csv dataframe:
write.csv(Gut_normalized_withoutBatch, 
          file= paste0(output_dir,"Gut_normalized_withoutBatch.csv")) 
# CHECK --> file save in output_dir

```

# Notes:

- Always check that output_dir is the correct destination before saving the file and check that export was successful.

- Check that counts dataframe (normalized without batch) has the correct columns.

- If dataframe look Ok, proceed to 03_DataWideToLong.qmd which is required for Boxplots.

- This dataframe is also the input for creating heatmaps.

```{r}
sessionInfo()
```
