---
title: "DataWideToLong"
author: "Nicolas Veland"
format: html
editor: visual
---

# Conversion of Count Dataframe Wide format to Long format

This script starts with the output data from **02_normalization-QC.qmd** script.

This script perform the following steps:

1.  Modify counts dataframe from wide to long.
2.  Reorganize dataframe by separating strings, creating new columns and renaming.
3.  Reorder columns and replace sample labels if needed.
4.  Export normalized count wide dataframe.

```{r}
# Load Libraries
#| echo: false
#| message: false
library(tidyverse)
```

Start with a normalized dataframe:

```{r}
# 1. Modifiy counts dataframe from wide to long:
Gut_long_vst_counts <- pivot_longer(Gut_normalized_withoutBatch, -c(Ensembl_id, gene), names_to = "Sample", values_to = "Expression")

# 2. Reorganize dataframe by separating strings, creating new columns and renaming:
Gut_long_vst_counts <- separate(Gut_long_vst_counts, Sample, into=c("Genotype", "Replicate"), sep = "_", remove = FALSE, extra = "merge")

#Epidermis_long_vst_counts <- separate(Epidermis_long_vst_counts, Replicate, into=c("Genotype", NA), sep = "_", remove = FALSE)

Gut_long_vst_counts <- separate(Gut_long_vst_counts, Replicate, into=c(NA, "Genotype"), sep = "_", remove = FALSE)

# 3. Reorder columns and replace sample labels if needed
Gut_long_vst_counts <- Gut_long_vst_counts[,c(1:3,5,4,6)]

Gut_long_vst_counts <- mutate(Gut_long_vst_counts, Genotype = str_replace(Genotype, "HET", "TCR KO"))

# 4. Exporting Normalized Long Counts as csv dataframe:
write.csv(as.data.frame(Gut_long_vst_counts), file= paste0(output_dir,"Gut_long_vst_counts.csv"))

```

## Notes:

- Always check that output_dir is the correct destination before saving the file and check that export was successful.

- Check that counts dataframe (long_vst_counts) has the correct columns.

- If dataframe look Ok, proceed to 05_Boxplots.qmd, but to include p values, 04_DiffGeneExpression.qmd script needs to be run first as p value included in boxplots comes from DESeq padj (p adjusted) values.

```{r}
sessionInfo()
```
