---
title: "DifferentialGeneExpression"
author: "Nicolas Veland"
format: html
editor: visual
---

# Differential Gene Expression Analysis

This script starts with the output of **01_processing.qmd** script.

This scripts perform all the steps required for Differential Gene Expression Analysis using DESeq package. It also contains the parameters to define DEG.

### A. DESeq2 analysis.

1.  Create a DESeqDataSet object with Batch effect correction.

2.  Define the reference level in DESeqDataSet object.

3.  Run DESeq analysis function.

4.  Examining DESeq results.

5.  Exporting DESeq2 results as csv dataframe.

### B. Read and organize DESeq2 analysis.

1.  Read DESEq results file as dataframe.

2.  Check all Ensembl_ids matching to DESeq dataframe

3.  Insert gene_name comlumn in DESeq dataframe.

### C. Define DEG from DESeq2 analysis.

1.  Filter by removing NA in padj and setting up baseMean.

2.  Setup significance cut-off.

3.  Count number of genes per condition and assigned to groups based on expression changes.

4.  Create Table with DEG numbers

5.  Create gene_change column

6.  Export complete DESeq dataframe as csv

```{r}
# Load Libraries
#| echo: false
#| message: false
library(DESeq2)
library(tidyverse)
library(magrittr)
```

## A. DESeq2 analysis

```{r}
######### A. DESeq2 analysis.-----------------------------------

# 1. Create a DESeqDataSet object with Batch effect correction:
# design= ~ batch + condition --> The design indicates how to model the samples, here, that we want to measure the effect of the condition, controlling for batch differences. The two factor variables batch and condition should be columns of coldata.
# Note: In order to benefit from the default settings of the package, you should put the variable of interest at the end of the formula and make sure the control level is the first level.
dds_Tissue = DESeqDataSetFromMatrix(
    countData = Gut_Raw_Counts,
    colData = Gut_colData,
    design= ~ Batch + Genotype)
```

```{r}
# 2. Define the reference level in DESeqDataSet object:
#levels(dds$Genotype) # this prints the order of the levels which is alphabetically ordered by default. DESeq treats first level as reference, so alphabetically order will compare KO vs WT using KO as reference.

levels(dds_Tissue$Genotype)


# relevel() function to change order of levels using WT as first level to use it as reference to which other levels are compared against (i.e. KO vs WT)
dds_Tissue$Genotype = relevel(dds_Tissue$Genotype, "WT") 


# run again levels(dds$Genotype) to check the correct order 
levels(dds_Tissue$Genotype)

```

```{r}
# 3. Run DESeq analysis function:
Gut_DESeq = DESeq(dds_Tissue)

```

```{r}
# 4. Examining DESeq results:
resultsNames(Gut_DESeq)

Gut_DESeq = results(Gut_DESeq, name = "Genotype_KO_vs_WT")

# Results with lfcShrink: This is optional alternative approach for the analysis using DESeq2. For more details, please refer to the DESeq2 documentation.
#Gut_DESeq2 = lfcShrink(dds_Tissue, coef = "Genotype_KO_vs_WT", type = "apeglm")
```

```{r}
# 5. Exporting DESeq2 results as csv dataframe:
write.csv(as.data.frame(Gut_DESeq), file= paste0(output_dir,"Gut_DESeq.csv"))
```

## B. Read and organize DESeq2 analysis.

```{r}
######### B. Read and organize DESeq2 analysis.-----------------------------------

# 6. Read DESEq results file as dataframe:
Gut_DESeq <- read.csv(file= "~/Documents/My_R/Github_Cg1-KO/RNAseq/Figure_3/Gut_DESeq.csv")


# 7. Check all Ensembl_ids (column X) in Ensembl_symbol dataframe matches the Ensembl_ids (column X) in DESeq dataframe:
all(Gut_DESeq$X == geneid$Ensembl)

# 8. Insert gene_name comlumn in DESeq dataframe from Ensembl_symbol dataframe symbol column:
Gut_DESeq$gene_name = geneid$symbol


# Check that TCR genes are annotated correctly and double check with Ensembl ID (e.g. ENSMUSG00000076746 = Trgv6). If not present, may need to be added manually.

Gut_DESeq$gene_name[21741] = "Trgv6"

head(Gut_DESeq)




```

From this point, DESEQ2 dataframe has gene name to facilitate searching genes.

## C. Define DEG from DESeq2 analysis.

```{r}
######### C. Define DEG from DESeq2 analysis.-----------------------------------

# 9. Filter by removing NA in padj and setting up baseMean:
Gut_DESeq %<>%
  as.data.frame() %>%
  filter(!is.na(padj)) %>%
  filter(baseMean >= 10)

# Are there trascripts without a gene_name that have NA instead? which ones?
length(which(is.na(Gut_DESeq$gene_name)))


# Replace transcripts with NA in gene_name for their Ensembl ID:
Gut_DESeq$gene_name[which(is.na(Gut_DESeq$gene_name))] <- Gut_DESeq$X[which(is.na(Gut_DESeq$gene_name))]

# 10. Significance cut-off:
Gut_DESeq %<>%
  mutate(sig = if_else(padj <= 0.05, "significant", "not-significant"))

head(Gut_DESeq)
```

```{r}
# 11. Count number of genes per condition and assigned to groups based on expression changes

#Down_DEG
length(which(Gut_DESeq$sig == "significant" & Gut_DESeq$log2FoldChange < 0))
N_Down_DEG = length(which(Gut_DESeq$sig == "significant" &
                            Gut_DESeq$log2FoldChange < 0))
Down_DEG = which(Gut_DESeq$sig == "significant" & 
                   Gut_DESeq$log2FoldChange < 0)

#Up_DEG
length(which(Gut_DESeq$sig == "significant" & Gut_DESeq$log2FoldChange > 0))
N_Up_DEG = length(which(Gut_DESeq$sig == "significant" &
                          Gut_DESeq$log2FoldChange > 0))
Up_DEG = which(Gut_DESeq$sig == "significant" &
                 Gut_DESeq$log2FoldChange > 0)

#All_DEG:
length(which(Gut_DESeq$sig == "significant"))
N_All_DEG = length(which(Gut_DESeq$sig == "significant"))
All_DEG = which(Gut_DESeq$sig == "significant")

#Epi_DEG_df = Epi_res_noBatch_DESEq %>% filter(Epi_res_noBatch_DESEq$sig == "significant")
#Epi_DEG_list = Epi_DEG_df$gene_name

# Unchanged genes:
length(which(Gut_DESeq$sig == "not-significant"))
N_Unchanged = length(which(Gut_DESeq$sig == "not-significant"))

# Total genes:
length(Gut_DESeq$sig)
N_Total = length(Gut_DESeq$sig)

N_All_DEG == sum(N_Down_DEG, N_Up_DEG)
N_Total == sum(N_Down_DEG, N_Up_DEG, N_Unchanged)

# print numbers:
N_Down_DEG
N_Up_DEG 
N_All_DEG
N_Unchanged
N_Total
```

```{r}
# 12. Create Table with DEG numbers
# create matrix with 4 columns and 4 rows
table_DEG = matrix(c(1:5), ncol=5, byrow=TRUE)

# specify the column names and row names of matrix
colnames(table_DEG) = c("Down_DEG", "Up_DEG", "All_DEG", "Unchanged", "Total")
rownames(table_DEG) <- c("Gut_DESeq")

table_DEG[1,1] = N_Down_DEG
table_DEG[1,2] = N_Up_DEG
table_DEG[1,3] = N_All_DEG
table_DEG[1,4] = N_Unchanged
table_DEG[1,5] = N_Total

table_DEG

table_DEG = as.table(table_DEG)
#write.csv(table_DEG, file= paste0(output_dir,"Table_Gut_DESeq.csv"), row.names = TRUE, col.names = TRUE)

table_DEG
```

```{r}
# 13. Create gene_change column:
Gut_DESeq$gene_change = "Unchanged"

Gut_DESeq$gene_change[Down_DEG] = "Down in KO"
Gut_DESeq$gene_change[Up_DEG] = "Up in KO"

# 14. Export complete DESeq dataframe as csv:
write.csv(as.data.frame(Gut_DESeq), 
          file= paste0(output_dir,"Gut_DESeq.csv"))
```

# Notes:

- Always check that output_dir is the correct destination before saving the file and check that export was successful.

- Check that Gut_DESeq dataframe has the correct columns and information.

- If dataframe look Ok, proceed to plotting using 05_Boxplots.qmd, 06_Volcano.qmd and 07_GSEA.qmd.

### **From here, DESeq results are saved in output_dir and are ready for downstream analyses such as plotting and GSEA.**

```{r}
sessionInfo()
```
