---
title: "Boxplots"
author: "Nicolas Veland"
format: html
editor: visual
---

# Boxplots from Gene Expression Counts

- This script allows to create boxplots using ggplot2 package to visualize expression levels of specific genes using normalized counts on a dataframe that has been converted from wide to long and its the output of the script "**03_DataWideToLong.qmd**".

- The significance indicated by the p value in the boxplot is taken from the padj (p adjusted) value from dataframe DESeq that is the annotated output from DESeq, which was generated with the script "**04_DiffGeneExpression.qmd**" and relies on the values and comparisons selected during the analysis using DESeq2 package. For more detailed information, please refer to DESeq2 documentation.

```{r}
# Load Libraries
#| echo: false
#| message: false
library(ggplot2)
library(ggrepel)
library(tidyverse)
library(magrittr)
```

1.  The long count dataframe can be imported using the following code:

```{r}
# 1. Read CLEAN FINAL filtered count wide dataframe as csv:
Tissue_long_vst_counts <- read.csv(file= paste0(output_dir,"Tissue_long_vst_counts.csv"))
Tissue_long_vst_counts = Tissue_long_vst_counts[,-1] # to remove the first column X with row numbers
```

2.  The DESeq dataframe can be imported using this code

```{r}
# 2. Read DESEq results file as dataframe:
Tissue_DESeq <- read.csv(file= paste0(output_dir,"Tissue_DESeq.csv"))
Tissue_DESeq = Tissue_DESeq[,-1] # Optional to remove first column in case is numbers from row names. 
```

3.  Create a boxplot with one gene at the time.

```{r}
# 7. Plot normalized data for QC GOI using Boxplots with using vst counts:
# GOI = Trgc1 (the gene that is deleted in TCR KO samples)
mydata = Tissue_long_vst_counts
mydata_source = "Tissue_long_vst_counts"
mygene = "Trgc1"
mydata_pvalue = Tissue_DESeq

# Extract the p-value for the current gene
  pvalue <- mydata_pvalue[which(mydata_pvalue$gene_name == mygene), "padj"]
  pvalue
  
# Check if the pvalue is available
  if (is.na(pvalue)) {
    p_label <- "padj = NA"
  } else {
    p_label <- paste("p =", format(pvalue, digits = 2))
  }
p_label

  
ggplot(mydata %>% dplyr::filter(gene == mygene), aes(x = Genotype, y = 2^Expression)) + #2^ for reverse Log transformed (or vst transforemd) data to plot counts in high values
  geom_boxplot(aes(fill = Genotype)) +
  geom_jitter(width = 0.05, size = 3) +
  scale_x_discrete(limits=c("WT", "TCR KO")) +
  scale_fill_manual(values = c("salmon","grey40")) + # c("red","dodgerblue")
  labs(y = "Normalized Counts", title = paste("Tissue: expression of", mygene)) +
  geom_text_repel(aes(label = Replicate)) + #or label = Replicate depending on column name
  #theme_bw(base_size = 12) +
  theme_classic(base_size = 12) +
  #theme(panel.border = element_rect(colour = "black")) +
  theme(axis.text = element_text(colour = "black", size = 12)) +
  theme(axis.text.x = element_text(angle = 0, colour = "black", size = 12)) +
  theme(legend.position="none") +
  # Add the p-value annotation
  #annotate("text", x = 1.5, y = max((subset(mydata, gene == mygene))$Expression) * 1.02, label = p_label, size = 5) 
  annotate("text", x = 1.5, y = (2^max((subset(mydata, gene == mygene))$Expression) * 1.1), label = p_label, size = 5) # uncomment if using 2^Expression if using this in aes
```

```{r}
# Save boxplot with ggsave: CAREFULLY RENAME
ggsave(paste0(output_dir,"Bp_unlog_", mygene, "_", mydata_source, ".pdf"),  width = 3.5, height = 3.5, units = "in")
```

4.  Define GOI list

```{r}
GOI <- c("Trgc1", "Trgv5", "Trgc3")
```

5.  Make Boxplots in Loop

```{r}
# 5. Boxplots of multiple genes in Loop saved in the same pdf,
mydata = Tissue_long_vst_counts
mydata_source = "Tissue_long_vst_counts"
mydata_pvalue = Tissue_DESeq


pdf(paste0(output_dir,"BP_TCR-genes_unlog", mydata_source, ".pdf"), width = 3.5, height = 3.5)
for (mygene in GOI){
  print(mygene)
 
  if (!mygene %in% mydata_pvalue$gene_name) {
    #pvalue = "NA"
    p_label <- "p = NA"
  } else {
     # Extract the p-value for the current gene
  pvalue <- mydata_pvalue[which(mydata_pvalue$gene_name == mygene), "padj"]
  p_label <- paste("p =", format(pvalue, digits = 2))
  }
  
  bp<- ggplot(mydata %>% dplyr::filter(gene == mygene), aes(x = Genotype, y = 2^Expression)) + #2^ for reverse Log transformed (or vst transforemd) data to plot counts in high values (y = 2^Expression)
  geom_boxplot(aes(fill = Genotype)) +
  geom_jitter(width = 0.05, size = 3) +
  scale_x_discrete(limits=c("WT", "TCR KO")) +
  scale_fill_manual(values = c("salmon","grey40")) +
  labs(y = "Normalized Counts", title = paste("Tissue (Vg5):", mygene)) +
  geom_text_repel(aes(label = Replicate)) + #or label = Replicate depending on column name
  #theme_bw(base_size = 12) +
  theme_classic(base_size = 14) +
  #theme(panel.border = element_rect(colour = "black")) +
  theme(axis.text = element_text(colour = "black", size = 14)) +
  theme(axis.text.x = element_text(angle = 0, colour = "black", size = 14)) +
   theme(legend.position="none") +
  # Add the p-value annotation
  annotate("text", x = 1.5, y = (2^max((subset(mydata, gene == mygene))$Expression) * 1.2), label = p_label, size = 5) # add 2^max if using this in aes
  
  print(bp)
  
}

dev.off()


```

# Notes:

- Always check that output_dir is the correct destination before saving the file and check that export was successful.

- Customizing different sections of the plot is possible. Please refer to ggplot documentation for additional details.
