---
title: "GSEA"
author: "Nicolas Veland"
format: html
editor: visual
---

# Gene Set Enrichement Analysis (GSEA)

- This script allows to perform a GSEA using fgsea package to visualize gene expression changes globally across reference gene sets.

- Starts with a dataframe that is the annotated output from DESeq, which was generated with the script "**04_DiffGeneExpression.qmd**" and relies on the values and comparisons selected during the analysis using DESeq2 package. For more detailed information, please refer to DESeq2 documentation.

```{r}
# Load Libraries
#| echo: false
#| message: false
library(fgsea)
library(msigdbr)
library(tidyverse)
library(dplyr)
library(magrittr)
library(ggplot2)
library(ggrepel)
library(clusterProfiler)
library(enrichplot)
library(patchwork)
library(ggplotify)

#library(genekitr)
#library(msigdb)

```

1.  Read DESEq results file as dataframe:

```{r}
# 1. Read DESEq results file as dataframe:
Dermis_DESeq <- read.csv(file= "~/Documents/My_R/Github_Cg1-KO/RNAseq/Figure_2/Dermis_DESeq.csv")
Dermis_DESeq = Dermis_DESeq[,-1] # Optional to remove first column in case is numbers from row names. 

```

2.  Prepare ranked list of genes

```{r}
# 2. Prepare ranked list of genes:

# Try Ranking option 1: using log2FC*log10(pvalue)
# The magnitude of the ranking will be given by the p value (since use -log10(), the more significant a gene is differentially expressed, the larger that value will be). The sign of the ranking (positive or negative) will be given by the fold-change: upregulated genes will have positive rankings, downregulated genes will have negative rankings.
gene.list <- Dermis_DESeq[!grepl("ENSMUS",Dermis_DESeq$gene_name),] 


gene.rank <- sign(gene.list$log2FoldChange)*(-log10(gene.list$pvalue)) 
# we will use the signed p values from spatial DGE as ranking

names(gene.rank) <- gene.list$gene_name # genes as names#

#names(gene.rank) <- Dermis_DESeq$X #replace gene_name in gene.rank by using Ensembl IDs from column X in case there are duplicates

###----

#head(ranked_genes)

head(gene.rank)

# Plot ranking to check max and min values:
#ranked_genes <- sort(ranked_genes, decreasing = TRUE) # sort genes by ranking
#plot(ranked_genes)
gene.rank <- sort(gene.rank, decreasing = TRUE) # sort genes by ranking
plot(gene.rank)

```

3.  Find duplications and correct them.

```{r}
# Find duplications names, inspect them and fix the issue (if any)
any(duplicated(names(gene.rank)))
```

```{r}
# Inspect them and fix the issue (if any)
# In case there is duplications of genes names due to issues with gene mapping database (usually happens for lncRNA genes or predicted genes). Is corrected manually by: 
# 1. find the duplicated gene name in counts dataframe; 

# 2. then search on Ensemble web using the Ensembl_id and identify which is the real and which is the incorrect; 
Dermis_DESeq %>%
  mutate(row = row_number()) %>%
  group_by(gene_name) %>%
  filter(n() > 1) %>%
  arrange(gene_name, desc(abs(log2FoldChange))) %>%
  dplyr::select(row, gene_name, X, sig)
```

```{r}
# 3. use the [] indexing with the row number and replace $gene with the "corrected_name" as shown in example. Otherwise next step gives Error.
Dermis_DESeq$gene_name[3586] = "Naa12" 

```

```{r}
# Check duplications are no longer present
Dermis_DESeq %>%
  mutate(row = row_number()) %>%
  group_by(gene_name) %>%
  filter(n() > 1) %>%
  arrange(gene_name, desc(abs(log2FoldChange))) %>%
  dplyr::select(row, gene_name, X, sig)

```

```{r}
# Re-write fixed dataframe (without duplicates)
write.csv(Dermis_DESeq, file = "~/Documents/My_R/Github_Cg1-KO/RNAseq/Figure_2/Dermis_DESeq.csv")
```

```{r}
# Check max and min ranking values:
# fgsea() does not accept non-finite numbers as rankings
max(gene.rank)
min(gene.rank)
```

4.  Prepare gene sets/pathways:

```{r}
# 4. Prepare gene sets/pathways:

# 4.1 Retrieve msigdb gene set/pathway data and generate dataframe 
#category = "H" #<---- hallmark_gene_sets
#category = "C7" #<---- immuno_gene_sets 
#category = "C2", subcategory = "CP:REACTOME"
#category = "C2", subcategory = "CP:KEGG"

msigdbr_df <- msigdbr(species = "Mus musculus", category = "H") #

# 4.2 Format msigdbr_df for fgsea
gene_set = split(x = msigdbr_df$gene_symbol, f= msigdbr_df$gs_name)

gene_set_name = "Hallmarks" #######<--- CHANGE CORRECT PATHWAY NAME

head(gene_set)

 
# Optional: 2.3 Format geneset for clusterProfiler:
#geneset_H <- msigdbr(species = "Mus musculus", category = "H") %>% 
#  dplyr::select(gs_name, gene_symbol)
#head(geneset_H)
#gene_set_name = "Hallmarks"
#
#geneset_IMM <- msigdbr(species = "Mus musculus", category = "C7") %>% 
#  dplyr::select(gs_name, gene_symbol) 
#head(geneset_IMM)
```

5.  Run GSEA using fgsea package

```{r}
# 5. Run GSEA using fgsea package:
GSEA_HM <- fgsea(pathways = gene_set, # List of gene sets to check
                 stats = gene.rank,
                 scoreType = 'std', # in this case we have both pos and neg rankings. if only pos or neg, set to 'pos', 'neg'
                 minSize=10, maxSize = 500)

head(GSEA_HM)

GSEA_HM

head(GSEA_HM[order(padj, -abs(NES)), ], n=10)

```

6.  Inspect and subset Top GSEA results

```{r}
# 6. Inspect and subset Top GSEA results
sum(GSEA_HM[, padj < 0.05])

length(GSEA_HM$pathway)
head(GSEA_HM, 20)
head(GSEA_HM %>% filter(padj <= 0.05))

Top5 = head(GSEA_HM %>%  arrange(-abs(NES)), n = 5) %>% arrange(desc(NES)) 
#Top5 = head(GSEA_HM %>% filter(padj <= 0.05) %>%  arrange(-abs(NES)), n = 5) %>% arrange(desc(NES)) 
Top5


```

```{r}
# Inspect top GSEA results
range(Top5$size)
min_size <- min(Top5$size)
max_size <- max(Top5$size)
mid_size <- round(mean(range(Top5$size)))

min_size
mid_size
max_size

Top5 = as.data.frame(Top5)
Top5 = Top5[, -8]
write.csv(Top5, file = paste0(output_dir,"Dermis_GSEA-Top5.csv"))

#as.table(head(GSEA_Result %>%  arrange(-abs(NES)), n = 5) %>% arrange(desc(NES))) 
```

```{r}
# Read GSEA results file as dataframe:
Top5 <- read.csv(file = paste0(output_dir,"Dermis_GSEA-Top5.csv"))
Top5 = Top5[,-1] # Optional to remove first column in case is numbers from row names. 
```

7.  Plot Summary of GSEA results as a Dotplot

```{r}
# 7.  Plot summary of top pathways: Dotplot ***********************
Top5 %>%
  ggplot(aes(x = reorder(pathway, NES), y = NES, color = padj, size = size)) +
  geom_point() +
  scale_color_gradient(low = "red", high = "blue", guide = guide_colourbar(reverse = TRUE)) +
  scale_size(
  range = c(2, 5),
  breaks = c(134, 167, 196),  # set manually if preferred
  name = "Gene count"
) +
  coord_flip() +
  xlab("Pathways") +
  ylab("Enrichment Score (NES)") +
  ggtitle(paste0(" Top GSEA Dermis ", gene_set_name)) +
  guides(
    color = guide_colourbar(order = 1, reverse = TRUE),
    size  = guide_legend(order = 2)) +
  theme_bw() +
  theme(plot.title = element_text(size = 16)) +
  theme(plot.colour = element_text(size = 26)) +
  theme(axis.text.x = element_text(angle = 0, colour = "black", size = 12)) +
  theme(axis.text.y = element_text(angle = 0, colour = "black", size = 10)) +
  theme(legend.title = element_text(size = 12)) +
  theme(legend.text = element_text(size = 10))
```

```{r}
# Save GSEA summary table plot:
outTitle <- paste0("GSEA_DP_Dermis_", gene_set_name)
ggsave(paste0(output_dir, outTitle, ".pdf"),  width = 6.5, height = 4, units = "in", dpi = 72)
```

```{r}
#  GSEA plot the most significantly enriched pathway
plotEnrichment(gene_set[[head(GSEA_HM[order(padj), ], 1)$pathway]],
               gene.rank) + 
  labs(title = head(GSEA_HM[order(padj), ], 1)$pathway)
```

```{r}
TopSinglePathway = head(GSEA_HM[order(padj), ], 1)$pathway

p =plotEnrichment(gene_set[[TopSinglePathway]],
               gene.rank) +
  labs(title = paste0("Top GSEA pathway in Dermis: \n", TopSinglePathway)) +
  scale_x_continuous('Rank') +
  scale_y_continuous('Enrichment score (ES)')
p + theme(plot.title = element_text(size=15))

```

8.  GSEA using clusterProfiler package (for plotting of enrichment score across genesets)

```{r}
# Format geneset for clusterProfiler:
geneset_H <- msigdbr(species = "Mus musculus", category = "H") %>% 
  dplyr::select(gs_name, gene_symbol)
head(geneset_H)
gene_set_name = "Hallmarks"
```

```{r}
# 8. ClusterProfiler GSEA analysis

#GSEA analysis requires a ranked gene list format, which contains three features:
  # 1. numeric vector: fold change or other type of numerical variable
  # 2. named vector: every number has a name, the corresponding gene ID
  # 3. sorted vector: number should be sorted in decreasing order
#You can prepare your own geneList via the following command:

## feature 1: numeric vector
gene.list2 <- Dermis_DESeq[!grepl("ENSMUS",Dermis_DESeq$gene_name),] 

geneList <- sign(gene.list2$log2FoldChange)*(-log10(gene.list2$pvalue)) # we will use the signed p values from spatial DGE as ranking

## feature 2: named vector
names(geneList) = as.character(gene.list2$gene_name)

## feature 3: decreasing order
geneList = sort(geneList, decreasing = TRUE)
head(geneList)

# Run GSEA using ClusterProfiler GSEA function
GSEA_CP <- GSEA(geneList, exponent = 0, pvalueCutoff = 10, TERM2GENE = geneset_H) #<- ISSUE HERE, NOT WORKING :S
head(GSEA_CP)
```

9.  Visualize the distribution of the gene set and the enrichment score (GSEAplot) with gseaplot2 function from enrichplot package

```{r}
# 9. Visualize the distribution of the gene set and the enrichment score with enrichplot

gseaplot2(GSEA_CP, geneSetID = 1, title = paste0("Top GSEA pathway in Dermis: \n", GSEA_CP$Description[1]), pvalue_table = TRUE)
```

```{r}
# Customize function gseaplot2 from enrichplot: to add NES value and edit some aesthetics:
gseaplot2_new <- enrichplot:::gseaplot2

fix(gseaplot2_new)
#trace("gseaplot2_new", edit = TRUE)

#In the function, look for line 79: "if (pvalue_table) {..." and modify the code from lines 80, 81, 82 and 90 to look like this:
#        pd <- x[geneSetID, c("NES", "p.adjust")]
#        rownames(pd) <- ""
#        pd <- pd[, ]
#        if (is.null(pvalue_table_rownames)) {
#            rows <- NULL
#        }
#        else {
#            if (length(pvalue_table_rownames) != 1) {
#                stop("the length of `pvalue_table_rownames` should be equal to 1")
#            }
#            rows <- x[geneSetID, ""]

# Also to modify aesthetics: modify line 12 and replace panel.grid.major.y wtih:
#       panel.grid.major.y = element_line(colour = "grey92"), panel.grid.minor.y = element_blank()) + 

# Finally, change color of lower plot from "grey92" to "black" by modifying line 62.

# Check if plot looks as desired by running this code:
gseaplot2_new(GSEA_CP, geneSetID = 1, title = paste0("Top GSEA pathway in Dermis: \n", GSEA_CP$Description[1]), pvalue_table = TRUE)
  
  
```

```{r}
# Create custome GSEA score plot for Top GSEa pathway
gseaplot2_new(GSEA_CP, geneSetID = 1, title = paste0("Top GSEA pathway in Dermis: \n", GSEA_CP$Description[1]), pvalue_table = TRUE)
```

```{r}
# Save gseaplot:
outTitle <- paste0("GSEA_Test_", pathway_name, TopSinglePathway)
ggsave(paste0(output_dir, outTitle, ".pdf"),  width = 6, height = 4, units = "in", dpi = 72)
```

10. Plot multiple TOP GSEA pathways from enrichplot using ggplotify + patchwork

```{r}
# 10. Plot multiple TOP GSEA pathways from clusterProfiler using ggplotify + patchwork
p1 = as.ggplot(gseaplot2_new(GSEA_CP, geneSetID = 1, title = GSEA_CP$Description[1], pvalue_table = TRUE))
p2 = as.ggplot(gseaplot2_new(GSEA_CP, geneSetID = 2, title = GSEA_CP$Description[2], pvalue_table = TRUE))
p3 = as.ggplot(gseaplot2_new(GSEA_CP, geneSetID = 3, title = GSEA_CP$Description[3], pvalue_table = TRUE))
p4 = as.ggplot(gseaplot2_new(GSEA_CP, geneSetID = 4, title = GSEA_CP$Description[4], pvalue_table = TRUE))
  
patchwork <- (p1 + p2 + p3 + p4)
patchwork + plot_annotation(
  title = "Dermis",
  theme = theme(plot.title = element_text(size = 20)))

```

```{r}
# Save GSEA summary table plot: 
outTitle <- paste0("GSEA_CP_", gene_set_name , "_TopPathways")
ggsave(paste0(output_dir, outTitle, ".pdf"),  width = 9, height = 8.5, units = "in", dpi = 72)
```

# Notes:

- Always check that output_dir is the correct destination before saving the file and check that export was successful.

- Customizing different sections of the plot is possible. Please refer to fgsea, clusterProfiler and enrichplot documentation for additional details.
