---
title: "Heatmap"
author: "Nicolas Veland"
format: html
editor: visual
---

# Heatmap from RNAseq normalized counts

- This script allows to create a heatmap using pheatmap package to visualize gene expression changes across samples at a global scale visualization or for a specific set of genes of interests.

- Starts with a dataframe of normalized counts (vst nornalization) that where gene names were inserted as a column and is the annotated output from "**02_normalization-QC.qmd**" script.

- If a heatmap showing all DEG is needed, then a DEG dataframe needs to be subset and gene names mapped to counts dataframe to select only those genes. For this purpose, the output of "**04_DiffGeneExpression.qmd**" script must be required.

```{r}
# Load Libraries
#| echo: false
#| message: false
library(ggrepel)
library(tidyverse)
library(magrittr)
library(pheatmap)
library(dendsort)
library(dichromat)
library(viridis)
library(colorRamps)
```

1.  The normalized counts data can be imported using the following code:

```{r}
# 1. Read CLEAN FINAL filtered count wide dataframe as csv:
Tissue_normalized_withoutBatch <- read.csv(file= paste0(output_dir,"Tissue_normalized_withoutBatch.csv"))
Tissue_normalized_withoutBatch = Tissue_normalized_withoutBatch[,-1] # to remove the first column X with row numbers

```

2.  Define specific inputs, sample orders and custome colors palettes

```{r}
# 2. Subset dataframe to define all DEG gene inputs
Tissue_DEG_df = Tissue_DESeq %>% filter(Tissue_DESeq$sig == "significant")
Tissue_DEG_list = Tissue_DEG_df$gene_name

Tissue_DEG_counts = filter(Tissue_normalized_withoutBatch, gene %in% Tissue_DEG_list)
rownames(Tissue_DEG_counts) = Tissue_DEG_counts$gene
head(Tissue_DEG_counts)

myGOI_hm = Tissue_DEG_counts
myGOI_name = "_DEG_"

# Define order of samples:
my_col_order <- c("Skin_WT_1",  "Skin_WT_2",  "Skin_WT_4",  "Skin_Het_1", "Skin_Het_2", "Skin_Het_3")

# Define custome color pallete
BuWhOrRed = colorRampPalette(c("blue", "white", "orangered1"))(256)
colorHM = BuWhOrRed

```

3.  Create all DEG heatmap using pheatmap package

```{r}
# 3. Plot heatmap using pheatmap package
# if needed check this: https://slowkow.com/notes/pheatmap-tutorial/
callback = function(hc, ...){dendsort(hc, isReverse=T)} #<- to re-order clustering
tissue = "Add Tissue Here"
outTitle <- paste0("HM_Test", tissue, myGOI_name)
save_hm = paste0(output_dir, outTitle, ".pdf")

pheatmap(myGOI_hm[, my_col_order], #myGOI_hm[, 2:7] <-- for individual replicates
         scale = "row", #<- replace "column" to "none" to display source vst values
         display_numbers = F, #<- FALSE = no numbers
         cluster_cols = T, 
         cluster_rows = T,
         #kmeans_k = 8,
         clustering_distance_rows = "euclidean",
         clustering_method = "complete", 
         clustering_callback = callback, #<- for re-order clustering
         treeheight_row = 0, 
         treeheight_col = 0,
         #annotation_row = my_gene_col,
         cutree_cols = 2, 
         #cutree_rows = 2, 
         border_color = FALSE, #number_color = "black",
         show_rownames = FALSE,
         #show_colnames     = FALSE,
         angle_col = "45",
         cellwidth = 20, #cellheight = 18, 
         color = colorHM, 
         #breaks = breaks,
         #rownames_style = {"fontsize": 10, "fontstyle": "italic", "ha": "left", "position": (-0.1, 0)},
         fontsize_row = 3, fontsize_col = 11,
         #width = 4, height = 5, filename = save_hm, # <---- UNCOMMENT TO SAVE PLOT
         main = "Tissue Test all DEG") #myGOI_name
```

4.  Define GOI list

```{r}
GOI <- c("Tigit", "Cd69", "Itgae", "Tnfrsf9", "Cd28", "Jun",	"Junb", "Xcl1", "Il13", "Klrb1a", "Klra5", "Klra7", "Klrc1", "Ccl5")

GOI_counts <- filter(Tissue_normalized_withoutBatch, gene %in% GOI)

myGOI_hm = GOI_counts
myGOI_name = "GOI_"

my_col_order <- c("Skin_WT_1",  "Skin_WT_2",  "Skin_WT_4",  "Skin_Het_1", "Skin_Het_2", "Skin_Het_3")

rownames(myGOI_hm) = myGOI_hm$gene

head(myGOI_hm)

```

5.  Create GOI heatmap using pheatmap package

```{r}
# 5. Plot heatmap using pheatmap package
# if needed check this: https://slowkow.com/notes/pheatmap-tutorial/

callback = function(hc, ...){dendsort(hc, isReverse=T)} #<- to re-order clustering
tissue = "Tissue"
outTitle <- paste0("HM-GOI_Test2_", myGOI_name, tissue)
save_hm = paste0(output_dir, outTitle, ".pdf")

pheatmap(myGOI_hm[, my_col_order], #myGOI_hm[, 2:7] <-- for individual replicates
         scale = "row", #<- replace "column" to "none" to display source vst values
         display_numbers = F, #<- FALSE = no numbers
         cluster_cols = T, 
         cluster_rows = T,
         #kmeans_k = 5,
         clustering_distance_rows = "euclidean",
         clustering_method = "complete", 
         clustering_callback = callback, #<- for re-order clustering
         treeheight_row = 0, 
         treeheight_col = 0,
         #annotation_row = my_gene_col,
         cutree_cols = 2, 
         cutree_rows = 2, 
         border_color = FALSE, #number_color = "black",
         #show_rownames = FALSE,
         #show_colnames     = FALSE,
         angle_col = "45",
         cellwidth = 20, 
         cellheight = 15, 
         color = colorHM, 
         #breaks = breaks,
         #rownames_style = {"fontsize": 10, "fontstyle": "italic", "ha": "left", "position": (-0.1, 0)},
         fontsize_row = 10, fontsize_col = 11,
         #width = 3.5, height = 3.5, filename = save_hm, # <---- UNCOMMENT TO SAVE PLOT
         main = "Add Tissue Here") #myGOI_name
```

# Notes:

- Always check that output_dir is the correct destination before saving the file and check that export was successful.

- Customizing different sections of the plot is possible. Please refer to pheatmap documentation for additional details.
