---
title: "Normalization and QC"
author: "Nicolas Veland"
format: html
editor: visual
---

## Normalization and QC of RNAseq counts

1.  Using DESeq2 to normalize raw counts data using vst.
2.  Correct for batch effect.
3.  QC data inspection by PCA
4.  Save normalized count data matrix
5.  Add gene information
6.  Export normalized count wide dataframe.

```{r}
#| echo: false
#| message: false
library(DESeq2)
library(limma)
library(org.Mm.eg.db)
```

```{r}
######### II. Normalize using DESeq2 ----------------------------------------------

# 1. Create a DESeqDataSet object:
# design= ~ batch + condition --> The design indicates how to model the samples, here, that we want to measure the effect of the condition, controlling for batch differences. The two factor variables batch and condition should be columns of coldata.
# Note: In order to benefit from the default settings of the package, you should put the variable of interest at the end of the formula and make sure the control level is the first level.
dds = DESeqDataSetFromMatrix(
    countData = Tissue_Raw_Counts,
    colData = TissueEpi_colData,
    design= ~ Genotype)
```

```{r}
#  Plot a PCA to check batch effect
vsd_with_batch <- vst(dds)
plotPCA(vsd_with_batch, "Batch")
PCAwithBatch <- plotPCA(vsd_with_batch, "Genotype")

KOvsWT_colors = c("dodgerblue", "red")

PCAwithBatch + 
  geom_text(aes(label = colnames(vsd_with_batch)), show.legend = FALSE, size = 5, vjust = "inward", hjust = "inward", position = position_nudge(y = 1.3)) +
  theme(text=element_text(size=15), legend.text=element_text(size=8)) +
  theme_bw(base_size = 15) +
  ggtitle("PCA with batch effect") +
  #theme(legend.position = c(0.1, 0.1)) +
  theme(legend.background=element_blank()) +
  scale_colour_manual(values= KOvsWT_colors,
                     name="Genotype",
                     breaks=c("WT", "KO"),
                     labels=c("WT", "KO")
                     )
```

```{r}
# 2. Remove Batch effect using limma package
vsd_without_batch = vst(dds)
assay(vsd_without_batch) <- limma::removeBatchEffect(assay(vsd_with_batch), vsd_with_batch$Batch)

```

```{r}
# 3. Plot a PCA for QC after removing batch effect 
plotPCA(vsd_without_batch, "Batch")

PCAwithoutBatch <- plotPCA(vsd_without_batch, "Genotype")
PCAwithoutBatch + 
  geom_text(aes(label = colnames(vsd_without_batch)), show.legend = FALSE, size = 5, vjust = "inward", hjust = "inward", position = position_nudge(x = 0, y = 2)) +
        theme(text=element_text(size=15), legend.text=element_text(size=8)) +
        theme_bw(base_size = 15) + coord_fixed(ratio = 0.75) +
        ggtitle("PCA without batch effect") +
        #theme(legend.position = c(0.1, 0.1)) +
        theme(legend.background=element_blank()) +
        scale_colour_manual(values= KOvsWT_colors,
                     name="Genotype",
                     breaks=c("WT", "KO"),
                     labels=c("WT", "KO")
                     )
```

```{r}
# Save PCA plot after removing batch effect
pdf(paste0(output_dir,"TEST_PCA_removedBatch.pdf"), 
    width = 7, height = 4, pointsize = 10)
print(PCAwithoutBatch + 
        geom_text(aes(label = colnames(vsd_without_batch)), show.legend = FALSE, size = 5,
                  vjust = "inward", hjust = "inward", 
                  position = position_nudge(x = 0, y = 1.2)) +
        theme(text=element_text(size=15), legend.text=element_text(size=8)) +
        theme_bw(base_size = 15) + coord_fixed(ratio = 0.75) +
        ggtitle("TEST PCA without batch effect") +
        #theme(legend.position = c(0.1, 0.1)) +
        theme(legend.background=element_blank()) +
        scale_colour_manual(values= KOvsWT_colors,
                     name="Genotype",
                     breaks=c("WT", "KO"),
                     labels=c("WT", "KO")
                     ))
dev.off()
```

```{r}
# 4. Write batch corrected normalized vst matrix for downstream plotting (e.g. for Boxplots or Heatmaps using vst counts):
Tissue_normalized_withoutBatch <- assay(vsd_without_batch)



# 5. Modify counts matrix to add gene names

# a) Create geneid dataframe to assign gene names to Ensemble_id
Tissue_normalized_withoutBatch = as.data.frame(Tissue_normalized_withoutBatch)
Tissue_normalized_withoutBatch <- tibble::rownames_to_column(Tissue_normalized_withoutBatch, "Ensembl_id")

geneid = data.frame(
  Ensembl = Tissue_normalized_withoutBatch$Ensembl_id,
  symbol = NA
)

# b) Get gene symbols from org.Mm.eg.db
geneid$symbol = mapIds(
  org.Mm.eg.db,
  keys = Tissue_normalized_withoutBatch$Ensembl_id,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

# c) Check all Ensembl_ids in geneid dataframe matches the gene_ids in counts dataframe
all(Tissue_normalized_withoutBatch$Ensembl_id == geneid$Ensembl
    )

# d) Insert gene name comlumn in counts dataframe from geneid dataframe symbol column
Tissue_normalized_withoutBatch$gene = geneid$symbol

# From this point, counts dataframe has gene name to facilitate searching genes #<-------


```

# From this point, counts dataframe has gene name to facilitate searching genes #\<-------

```{r}
# 6. Exporting Normalized Counts as csv dataframe:
write.csv(Tissue_normalized_withoutBatch, 
          file= paste0(output_dir,"Tissue_normalized_withoutBatch.csv")) 
# CHECK --> file save in output_dir

```

Check that counts dataframe (normalized without batch) has the correct columns.

If dataframe look Ok, proceed to Long-Data_Boxplot script
