---
title: "Figure_4a_Human"
author: "Nicolas Veland"
format: html
editor: visual
---

# Figure_4a_Human

- This script allows to re-create volcano plot from Figure_4a_Human using ggplot2 package to visualize differential gene expression analysis after running DESeq.

- Starts with a dataframe ("Human_DESeq.csv") that is the annotated output from DESeq, which was generated with the script "**04_DiffGeneExpression.qmd"** and relies on the values and comparisons selected during the analysis using DESeq2 package. For more detailed information, please refer to DESeq2 documentation.

```{r}
# Load Libraries
#| echo: false
#| message: false
library(ggplot2)
library(ggrepel)
library(tidyverse)
library(magrittr)
library(org.Hs.eg.db)
```

1.  The data from DESeq can be imported using the following code:

```{r}
# 1. Read DESEq results file as dataframe:
Human_DESeq <- read.csv(file= paste0(output_dir,"Human_DESeq.csv"))

gene_id.hs  = data.frame(
  Ensembl = Human_DESeq$X,
  symbol = NA
)

# 2. Get gene symbols from org.Mm.eg.db
gene_id.hs$symbol = mapIds(
  org.Hs.eg.db,
  keys = Human_DESeq$X,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

# 3. Check all Ensembl_ids in geneid dataframe matches the gene_ids in counts dataframe
all(Human_DESeq$X == gene_id.hs$Ensembl
    )

Human_DESeq$gene_name = gene_id.hs$symbol
```

2.  A set of genes of interest (GOI) can be created as vector and use to label genes in the volcano

```{r}
# 2. Creat GOI list as a vector :
GOI <- c("TRDV1", "TIGIT", "TNFRSF9", "HAVCR2", "XCL1", "IFNG", "GZMA", "GZMB", "GZMK", "PRF1", "LTB", "CRTAM", "SCART1", "KLRD1", "KLRK1", "FOXP3", "IL5", "IL9", "IL13", "IL17RB","CCR3")
```

3.  Plot Volcano

```{r}
# 3. Volcano plot with GOI labeled:
Human_DESeq %>% 
  ggplot(aes(x=log2FoldChange, y=-log10(padj))) +
  geom_point(alpha=0.75, size = 2, aes(colour=gene_change)) +
  scale_colour_manual(values = c("blue","grey","red")) +
  geom_hline(yintercept = -(log10(0.05)), linetype="dashed", colour="grey") +
  geom_text_repel(data = Human_DESeq %>% filter(gene_name %in% GOI) %>% filter (padj <= 0.05), aes(label=gene_name), size = 5, fontface = "italic", max.overlaps = Inf, 
                  box.padding = 0.75, force_pull = 0.5, force = 10) + # max.overlaps = 40, 
  ylim(0,70) + 
  xlim(-7.5,7.5) + 
  labs(colour = "") +
  ggtitle("Human blood") +
  theme_bw(base_size = 14) +
  theme(legend.position = "none") + #"bottom"
  theme(plot.title = element_text(size = 16)) +
  theme(plot.colour = element_text(size = 26)) +
  theme(axis.text.x = element_text(angle = 0, colour = "black", size = 12)) +
  theme(axis.text.y = element_text(angle = 0, colour = "black",  size = 12)) +
  theme(legend.title = element_text(size = 14)) +
  theme(legend.text = element_text(size = 12))
```

```{r}
# 4. Save Volcano plot: CHECK NAME CAREFULLY
output_dir <- "/Users/velandn/Documents/My_R/Github_Cg1-KO/RNAseq/Figure_4/"
outTitle <- paste0("Figure_4a_Human")
ggsave(paste0(output_dir, outTitle, ".pdf"),  width = 5, height = 5, units = "in", dpi = 72)
```

# Notes:

- Always check that output_dir is the correct destination before saving the file and check that export was successful.

- Customizing different sections of the plot is possible. Please refer to ggplot documentation for additional details.

```{r}
sessionInfo()
```
