---
title: "Extended_Data_3f_Dermis"
author: "Nicolas Veland"
format: html
editor: visual
---

# Extended_Data_3f_Dermis

- This script allows to re-create Extended_Data_3f_Dermis from GSEA using fgsea package to visualize gene expression changes globally across reference gene sets.

- Starts with a dataframe that is the GSEA output from fgsea, which was generated with the script "**07_GSEA.qmd**" and relies on the values and comparisons selected during the DESeq analysis using DESeq2 package with the script "**04_DiffGeneExpression.qmd**". For more detailed information, please refer to DESeq2 documentation.

```{r}
# Load Libraries
#| echo: false
#| message: false
library(fgsea)
library(msigdbr)
library(tidyverse)
library(dplyr)
library(magrittr)
library(ggplot2)
library(ggrepel)
library(clusterProfiler)
library(enrichplot)
library(patchwork)
library(ggplotify)

#library(genekitr)
#library(msigdb)

```

1.  Read GSEa results file as dataframe:

```{r}
# 1. Read GSEA results file as dataframe:
Top5 <- read.csv(file = paste0(output_dir,"Dermis_GSEA-Top5.csv"))
Top5 = Top5[,-1] # Optional to remove first column in case is numbers from row names. 
```

2.  Plot Summary of GSEA results as a Dotplot

```{r}
# 2.  Plot summary of top pathways: Dotplot ***********************
Top5 %>%
  ggplot(aes(x = reorder(pathway, NES), y = NES, color = padj, size = size)) +
  geom_point() +
  scale_color_gradient(low = "red", high = "blue", guide = guide_colourbar(reverse = TRUE)) +
  scale_size(
  range = c(2, 5),
  breaks = c(134, 167, 196),  # set manually if preferred
  name = "Gene count"
) +
  coord_flip() +
  xlab("Pathways") +
  ylab("Enrichment Score (NES)") +
  ggtitle(paste0(" Top GSEA Dermis ", gene_set_name)) +
  guides(
    color = guide_colourbar(order = 1, reverse = TRUE),
    size  = guide_legend(order = 2)) +
  theme_bw() +
  theme(plot.title = element_text(size = 16)) +
  theme(plot.colour = element_text(size = 26)) +
  theme(axis.text.x = element_text(angle = 0, colour = "black", size = 12)) +
  theme(axis.text.y = element_text(angle = 0, colour = "black", size = 10)) +
  theme(legend.title = element_text(size = 12)) +
  theme(legend.text = element_text(size = 10))
```

```{r}
# Save GSEA summary table plot:
outTitle <- paste0("GSEA_DP_Dermis_", gene_set_name)
ggsave(paste0(output_dir, outTitle, ".pdf"),  width = 6.5, height = 4, units = "in", dpi = 72)
```

# Notes:

- Always check that output_dir is the correct destination before saving the file and check that export was successful.

```{r}
sessionInfo()
```
