---
title: "Extended_Data_3h_Dermis"
author: "Nicolas Veland"
format: html
editor: visual
---

# Extended_Data_3h_Dermis

- This script allows to re-create Extended_Data_3h_Dermis from GSEA using clusterProfiler and enrichplot packages to visualize gene expression changes globally across reference gene sets.

- Starts with a dataframe that is the annotated output from DESeq, which was generated with the script "**04_DiffGeneExpression.qmd**" and relies on the values and comparisons selected during the analysis using DESeq2 package. For more detailed information, please refer to DESeq2 documentation.

```{r}
# Load Libraries
#| echo: false
#| message: false
library(fgsea)
library(msigdbr)
library(tidyverse)
library(dplyr)
library(magrittr)
library(ggplot2)
library(ggrepel)
library(clusterProfiler)
library(enrichplot)
library(patchwork)
library(ggplotify)

#library(genekitr)
#library(msigdb)

```

1.  Read DESEq results file as dataframe:

```{r}
# 1. Read DESEq results file as dataframe:
Dermis_DESeq <- read.csv(file= "~/Documents/My_R/Github_Cg1-KO/RNAseq/Figure_2/Dermis_DESeq.csv")
Dermis_DESeq = Dermis_DESeq[,-1] # Optional to remove first column in case is numbers from row names. 

```

2.  GSEA using clusterProfiler package (for plotting of enrichment score across genesets)

```{r}
# Format geneset for clusterProfiler:
geneset_H <- msigdbr(species = "Mus musculus", category = "H") %>% 
  dplyr::select(gs_name, gene_symbol)
head(geneset_H)
gene_set_name = "Hallmarks"
```

```{r}
# 8. ClusterProfiler GSEA analysis

#GSEA analysis requires a ranked gene list format, which contains three features:
  # 1. numeric vector: fold change or other type of numerical variable
  # 2. named vector: every number has a name, the corresponding gene ID
  # 3. sorted vector: number should be sorted in decreasing order
#You can prepare your own geneList via the following command:

## feature 1: numeric vector
gene.list2 <- Dermis_DESeq[!grepl("ENSMUS",Dermis_DESeq$gene_name),] 

geneList <- sign(gene.list2$log2FoldChange)*(-log10(gene.list2$pvalue)) # we will use the signed p values from spatial DGE as ranking

## feature 2: named vector
names(geneList) = as.character(gene.list2$gene_name)

## feature 3: decreasing order
geneList = sort(geneList, decreasing = TRUE)
head(geneList)

# Run GSEA using ClusterProfiler GSEA function
GSEA_CP <- GSEA(geneList, exponent = 0, pvalueCutoff = 10, TERM2GENE = geneset_H) 
head(GSEA_CP)
```

9.  Visualize the distribution of the gene set and the enrichment score (GSEAplot) with gseaplot2 function from enrichplot package

```{r}
# 9. Visualize the distribution of the gene set and the enrichment score with enrichplot

gseaplot2(GSEA_CP, geneSetID = 1, title = paste0("Top GSEA pathway in Dermis: \n", GSEA_CP$Description[1]), pvalue_table = TRUE)
```

```{r}
# Customize function gseaplot2 from enrichplot: to add NES value and edit some aesthetics:
gseaplot2_new <- enrichplot:::gseaplot2

fix(gseaplot2_new)
#trace("gseaplot2_new", edit = TRUE)

#In the function, look for line 79: "if (pvalue_table) {..." and modify the code from lines 80, 81, 82 and 90 to look like this:
#        pd <- x[geneSetID, c("NES", "p.adjust")]
#        rownames(pd) <- ""
#        pd <- pd[, ]
#        if (is.null(pvalue_table_rownames)) {
#            rows <- NULL
#        }
#        else {
#            if (length(pvalue_table_rownames) != 1) {
#                stop("the length of `pvalue_table_rownames` should be equal to 1")
#            }
#            rows <- x[geneSetID, ""]

# Also to modify aesthetics: modify line 12 and replace panel.grid.major.y wtih:
#       panel.grid.major.y = element_line(colour = "grey92"), panel.grid.minor.y = element_blank()) + 

# Finally, change color of lower plot from "grey92" to "black" by modifying line 62.

# Check if plot looks as desired by running this code:
gseaplot2_new(GSEA_CP, geneSetID = 1, title = paste0("Top GSEA pathway in Dermis: \n", GSEA_CP$Description[1]), pvalue_table = TRUE)
  
  
```

```{r}
# Create custome GSEA score plot for Top GSEa pathway
gseaplot2_new(GSEA_CP, geneSetID = 1, title = paste0("Top GSEA pathway in Dermis: \n", GSEA_CP$Description[1]), pvalue_table = TRUE)
```

```{r}
# Save gseaplot:
outTitle <- paste0("GSEA_Test_", pathway_name, TopSinglePathway)
ggsave(paste0(output_dir, outTitle, ".pdf"),  width = 6, height = 4, units = "in", dpi = 72)
```

10. Plot multiple TOP GSEA pathways from enrichplot using ggplotify + patchwork

```{r}
# 10. Plot multiple TOP GSEA pathways from clusterProfiler using ggplotify + patchwork
p1 = as.ggplot(gseaplot2_new(GSEA_CP, geneSetID = 1, title = GSEA_CP$Description[1], pvalue_table = TRUE))
p2 = as.ggplot(gseaplot2_new(GSEA_CP, geneSetID = 2, title = GSEA_CP$Description[2], pvalue_table = TRUE))
p3 = as.ggplot(gseaplot2_new(GSEA_CP, geneSetID = 3, title = GSEA_CP$Description[3], pvalue_table = TRUE))
p4 = as.ggplot(gseaplot2_new(GSEA_CP, geneSetID = 4, title = GSEA_CP$Description[4], pvalue_table = TRUE))
  
patchwork <- (p1 + p2 + p3 + p4)
patchwork + plot_annotation(
  title = "Dermis",
  theme = theme(plot.title = element_text(size = 20)))

```

```{r}
# Save GSEA summary table plot: 
outTitle <- paste0("GSEA_CP_Dermis_", gene_set_name , "_TopPathways")
ggsave(paste0(output_dir, outTitle, ".pdf"),  width = 9, height = 8.5, units = "in", dpi = 72)
```

# Notes:

- Always check that output_dir is the correct destination before saving the file and check that export was successful.

```{r}
sessionInfo()
```
