---
title: "Extended_Data_3i_Dermis"
author: "Nicolas Veland"
format: html
editor: visual
---

# Extended_Data_3i_Dermis

- This script allows to re-create the heatmap of Extended_Data_3i_Dermis using pheatmap package to visualize gene expression changes across samples at a global scale visualization or for a specific set of genes of interests.

- Starts with a dataframe of normalized counts (vst nornalization) that where gene names were inserted as a column and is the annotated output from "**02_normalization-QC.qmd**" script.

```{r}
# Load Libraries
#| echo: false
#| message: false
library(ggrepel)
library(tidyverse)
library(magrittr)
library(pheatmap)
library(dendsort)
library(dichromat)
library(viridis)
library(colorRamps)
```

1.  The normalized counts data can be imported using the following code:

```{r}
# 1. Read CLEAN FINAL filtered count wide dataframe as csv:
Dermis_normalized_withoutBatch <- read.csv(file= paste0(output_dir,"Dermis_normalized_withoutBatch.csv"))
Dermis_normalized_withoutBatch = Dermis_normalized_withoutBatch[,-1] # to remove the first column X with row numbers

```

2.  Define specific inputs, sample orders and custom colors palettes

```{r}
# 2. Subset dataframe to define all DEG gene inputs

GOI <- c("Tigit", "Cd69", "Itgae", "Tnfrsf9", "Cd28", "Jun",	"Junb", "Xcl1", "Il13", "Klrb1a", "Klra5", "Klra7", "Klrc1", "Ccl5")

GOI_counts <- filter(Dermis_normalized_withoutBatch, gene %in% GOI)

myGOI_hm = GOI_counts
myGOI_name = "_GOI_"

rownames(myGOI_hm) = myGOI_hm$gene

head(myGOI_hm)

# Define order of samples:
my_col_order <- c("Dermis_WT_1",  "Dermis_WT_2",  "Dermis_WT_4",  "Dermis_Het_1", "Dermis_Het_2", "Dermis_Het_4")

# Define custome color pallete
BuWhOrRed = colorRampPalette(c("blue", "white", "orangered1"))(256)
colorHM = BuWhOrRed

```

3.  Create GOI heatmap using pheatmap package

```{r}
# 3. Plot heatmap using pheatmap package
# if needed check this: https://slowkow.com/notes/pheatmap-tutorial/

callback = function(hc, ...){dendsort(hc, isReverse=T)} #<- to re-order clustering
Tissue = "Dermis"
outTitle <- paste0("HM_", Tissue, myGOI_name)
save_hm = paste0(output_dir, outTitle, ".pdf")

pheatmap(myGOI_hm[, my_col_order], #myGOI_hm[, 2:7] <-- for individual replicates
         scale = "row", #<- replace "column" to "none" to display source vst values
         display_numbers = F, #<- FALSE = no numbers
         cluster_cols = T, 
         cluster_rows = T,
         #kmeans_k = 5,
         clustering_distance_rows = "euclidean",
         clustering_method = "complete", 
         clustering_callback = callback, #<- for re-order clustering
         treeheight_row = 0, 
         treeheight_col = 0,
         #annotation_row = my_gene_col,
         cutree_cols = 2, 
         cutree_rows = 2, 
         border_color = FALSE, #number_color = "black",
         #show_rownames = FALSE,
         #show_colnames     = FALSE,
         angle_col = "45",
         cellwidth = 20, 
         cellheight = 15, 
         color = colorHM, 
         #breaks = breaks,
         fontsize_row = 10, fontsize_col = 11,
         width = 4, height = 4.2, filename = save_hm, # <---- UNCOMMENT TO SAVE PLOT
         main = "Dermis GOI") #myGOI_name
```

# Notes:

- Always check that output_dir is the correct destination before saving the file and check that export was successful.

```{r}
sessionInfo()
```
