---
title: "Extended_Data_3e_Dermis"
author: "Nicolas Veland"
format: html
editor: visual
---

# Extended_Data_3e_Dermis

- This script allows to re-create the heatmap of Extended_Data_3e_Dermis using pheatmap package to visualize gene expression changes across samples at a global scale visualization or for a specific set of genes of interests.

- Starts with a dataframe of normalized counts (vst nornalization) that where gene names were inserted as a column and is the annotated output from "**02_normalization-QC.qmd**" script.

- If a heatmap showing all DEG is needed, then a DEG dataframe needs to be subset and gene names mapped to counts dataframe to select only those genes. For this purpose, the output of "**04_DiffGeneExpression.qmd**" script must be required.

```{r}
# Load Libraries
#| echo: false
#| message: false
library(ggrepel)
library(tidyverse)
library(magrittr)
library(pheatmap)
library(dendsort)
library(dichromat)
library(viridis)
library(colorRamps)
library(dplyr)
```

1.  The normalized counts data can be imported using the following code:

```{r}
# 1. Read CLEAN FINAL filtered count wide dataframe as csv:
Dermis_normalized_withoutBatch <- read.csv(file= paste0(output_dir,"Dermis_normalized_withoutBatch.csv"))
Dermis_normalized_withoutBatch = Dermis_normalized_withoutBatch[,-1] # to remove the first column X with row numbers

#  Import DESeq dataframe as csv:
Dermis_DESeq <- read.csv(file= "~/Documents/My_R/Github_Cg1-KO/RNAseq/Figure_2/Dermis_DESeq.csv")
Dermis_DESeq = Dermis_DESeq[,-1] # to remove the first column X with row numbers
```

2.  Define specific inputs, sample orders and custome colors palettes

```{r}
# 2. Subset dataframe to define all DEG gene inputs
Dermis_DEG_df = Dermis_DESeq %>% filter(Dermis_DESeq$sig == "significant")
Dermis_DEG_list = Dermis_DEG_df$gene_name

Dermis_DEG_counts = filter(Dermis_normalized_withoutBatch, gene %in% Dermis_DEG_list)

# Inspect for duplicated gene names and fix the issue (if any)
# In case there is duplications of genes names due to issues with gene mapping database (usually happens for lncRNA genes or predicted genes). This will cause an error for rownames assigning. Is corrected manually by: 
# find the duplicated gene name in counts dataframe; 
Dermis_DEG_counts %>%
  mutate(row = row_number()) %>%
  group_by(gene) %>%
  filter(n() > 1) %>%
  arrange(gene, row) %>%
  ungroup() %>%
  dplyr::select(row, gene, Ensembl_id)

#  use the [] indexing with the row number and replace $gene with the "corrected_name" as shown in example. Otherwise next step gives Error.
Dermis_DEG_counts$gene[1260] = "Naa12" 

rownames(Dermis_DEG_counts) = Dermis_DEG_counts$gene
head(Dermis_DEG_counts)

myGOI_hm = Dermis_DEG_counts
myGOI_name = "_DEG_"

# Define order of samples:
my_col_order <- c("Dermis_WT_1",  "Dermis_WT_2",  "Dermis_WT_4",  "Dermis_Het_1", "Dermis_Het_2", "Dermis_Het_4")

# Define custome color pallete
BuWhOrRed = colorRampPalette(c("blue", "white", "orangered1"))(256)
colorHM = BuWhOrRed

```

3.  Create all DEG heatmap using pheatmap package

```{r}
# 3. Plot heatmap using pheatmap package
# if needed check this: https://slowkow.com/notes/pheatmap-tutorial/
callback = function(hc, ...){dendsort(hc, isReverse=T)} #<- to re-order clustering
Tissue = "Dermis"
outTitle <- paste0("HM_", Tissue, myGOI_name)
save_hm = paste0(output_dir, outTitle, ".pdf")

pheatmap(myGOI_hm[, my_col_order], #myGOI_hm[, 2:7] <-- for individual replicates
         scale = "row", #<- replace "column" to "none" to display source vst values
         display_numbers = F, #<- FALSE = no numbers
         cluster_cols = T, 
         cluster_rows = T,
         #kmeans_k = 8,
         clustering_distance_rows = "euclidean",
         clustering_method = "complete", 
         clustering_callback = callback, #<- for re-order clustering
         treeheight_row = 0, 
         treeheight_col = 0,
         #annotation_row = my_gene_col,
         cutree_cols = 2, 
         #cutree_rows = 2, 
         border_color = FALSE, #number_color = "black",
         show_rownames = FALSE,
         #show_colnames     = FALSE,
         angle_col = "45",
         cellwidth = 20, #cellheight = 18, 
         color = colorHM, 
         #breaks = breaks,
         #rownames_style = {"fontsize": 10, "fontstyle": "italic", "ha": "left", "position": (-0.1, 0)},
         fontsize_row = 3, fontsize_col = 11,
         width = 4, height = 5, filename = save_hm, # <---- UNCOMMENT TO SAVE PLOT
         main = "Dermis all DEG") #myGOI_name
```

# Notes:

- Always check that output_dir is the correct destination before saving the file and check that export was successful.

```{r}
sessionInfo()
```
