---
title: "Figure_2g_Dermis"
author: "Nicolas Veland"
format: html
editor: visual
---

# Figure_2g_Dermis

- This script allows to re-create volcano plot from Figure_2g_Dermis using ggplot2 package to visualize differential gene expression analysis after running DESeq.

- Starts with a dataframe ("Dermis_DESeq.csv") that is the annotated output from DESeq, which was generated with the script "**04_DiffGeneExpression.qmd"** and relies on the values and comparisons selected during the analysis using DESeq2 package. For more detailed information, please refer to DESeq2 documentation.

```{r}
# Load Libraries
#| echo: false
#| message: false
library(ggplot2)
library(ggrepel)
library(tidyverse)
library(magrittr)
```

1.  The data from DESeq can be imported using the following code:

```{r}
# 1. Read DESEq results file as dataframe:
Dermis_DESeq <- read.csv(file= paste0(output_dir,"Dermis_DESeq.csv"))
Dermis_DESeq = Dermis_DESeq[,-1] # Optional to remove first column in case is numbers from row names. 
```

2.  A set of genes of interest (GOI) can be created as vector and use to label genes in the volcano

```{r}
# 2. Creat GOI list as a vector :
GOI <- c("Rasa2", "Itgae", "Jun",	"Junb", "Xcl1",	"Tigit", "Trgc1", "Trgc3", "Trgv4", "Trgv6", "Ccl3", "Ccl5", "Il13", "Tnfrsf9", "Klra1", "Klra3", "Klra5", "Klra6", "Klra7", "Klrb1", "Klrb1a", "Klrc1", "Klre1", "Klrg1", "Klrh1", "Klri1", "Klri2", "Klrk1", "Gzmb", "Gzma")
```

3.  Plot Volcano

```{r}
# 3. Volcano plot with GOI labeled:
Dermis_DESeq %>% 
  ggplot(aes(x=log2FoldChange, y=-log10(padj))) +
  geom_point(alpha=0.7, aes(colour=gene_change)) +
  scale_colour_manual(values = c("blue","grey","red")) +
  geom_hline(yintercept = -(log10(0.05)), linetype="dashed", colour="grey") +
  geom_text_repel(data = Dermis_DESeq %>% 
                    filter(gene_name %in% GOI) 
                  %>%  filter(padj <= 0.05), 
                  aes(label=gene_name), size = 5, fontface = "italic", max.overlaps = 20, 
                  force_pull = 0.1,  box.padding = 1, force = 6,) +  
  ylim(0,200) + 
  xlim(-8, 8) + 
  labs(colour = "") +
  ggtitle("Dermis") +
  theme_bw(base_size = 12) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(size = 14)) +
  theme(axis.text.x = element_text(angle = 0, colour = "black", size = 12)) +
  theme(axis.text.y = element_text(angle = 0, colour = "black",  size = 12)) +
  theme(legend.title = element_text(size = 14)) +
  theme(legend.text = element_text(size = 12))
```

```{r}
# 4. Save Volcano plot: CHECK NAME CAREFULLY
outTitle <- paste0("Figure_2g_Dermis")
ggsave(paste0(output_dir, outTitle, ".pdf"),  width = 5, height = 5, units = "in", dpi = 72)
```

# Notes:

- Always check that output_dir is the correct destination before saving the file and check that export was successful.

- Customizing different sections of the plot is possible. Please refer to ggplot documentation for additional details.

```{r}
sessionInfo()
```
