---
title: "Extended_Data_5a_Epidermis"
format: html
editor: visual
---

# Extended_Data_5b_Gut

This script re-creates the volcano plot from Extended_Data_5b_Gut with the output data generated from **1_consensus_peaks.sh and 2_downstream_DA_and_annotation.R** scripts.

```{r}

# Load Libraries
#| echo: false
#| message: false

library(tidyverse)
library(magrittr)
library(ggplot2)
library(ggrepel)
library(ggpubr)

```

```{r}
# Set wd:

setwd("~/Documents/My_R/ATAC")

# Create output_dir:
output_dir <- "/Users/velandn/Documents/My_R/Github_TEST/ATAC/"

ifelse(!dir.exists(file.path(output_dir)),
        dir.create(file.path(output_dir)),
        "Directory Exists")
```

A. Load files:

```{r}
# 1. Read CLEAN DA_K0_vs_WT.tsv as dataframe:
DA_Gut_KO_WT=read.delim(file = "~/Documents/My_R/Github_Cg1-KO/ATACseq/Gut_ATACseq/Gut_DA_KO_vs_WT.tsv")
```

B. Obtain counts for number of peaks and setup thresholds for significance and direction:

```{r}
# 1. Count total peaks
length(DA_Gut_KO_WT$Start)


# 2. Significance cut-off:
DA_Gut_KO_WT %<>%
  mutate(sig = if_else(FDR <= 0.05, "significant", "not-significant"))

# 3. Count number of peaks per group
# Total peaks:
length(DA_Gut_KO_WT$sig)
N_Total = length(DA_Gut_KO_WT$sig)

#All_DAR:
length(which(DA_Gut_KO_WT$sig == "significant"))
N_All_DAR = length(which(DA_Gut_KO_WT$sig == "significant"))
All_DAR = which(DA_Gut_KO_WT$sig == "significant")

#Down_DAR
length(which(DA_Gut_KO_WT$sig == "significant" & DA_Gut_KO_WT$logFC < 0))
N_Down_DAR = length(which(DA_Gut_KO_WT$sig == "significant" & DA_Gut_KO_WT$logFC < 0))
Down_DAR = which(DA_Gut_KO_WT$sig == "significant" & DA_Gut_KO_WT$logFC < 0)

#Up_DAR
length(which(DA_Gut_KO_WT$sig == "significant" & DA_Gut_KO_WT$logFC > 0))
N_Up_DAR = length(which(DA_Gut_KO_WT$sig == "significant" & DA_Gut_KO_WT$logFC > 0))
Up_DAR = which(DA_Gut_KO_WT$sig == "significant" & DA_Gut_KO_WT$logFC > 0)

# Unchanged DAR:
length(which(DA_Gut_KO_WT$sig == "not-significant"))
N_Unchanged_DAR = length(which(DA_Gut_KO_WT$sig == "not-significant"))

N_All_DAR == sum(N_Down_DAR, N_Up_DAR)
N_Total == sum(N_Down_DAR, N_Up_DAR, N_Unchanged_DAR)

```

```{r}
# print numbers:
N_Down_DAR
N_Up_DAR 
N_All_DAR
N_Unchanged_DAR
N_Total
```

```{r}
# 4. Creat Table with DAR numbers
# create matrix with 4 columns and 4 rows
table_DAR = matrix(c(1:5), ncol=5, byrow=TRUE)

# specify the column names and row names of matrix
colnames(table_DAR) = c("Down_DAR", "Up_DAR", "All_DAR", "Unchanged", "Total")
rownames(table_DAR) <- c("DA_Gut_KO_WT")

table_DAR[1,1] = N_Down_DAR
table_DAR[1,2] = N_Up_DAR
table_DAR[1,3] = N_All_DAR
table_DAR[1,4] = N_Unchanged_DAR
table_DAR[1,5] = N_Total

table_DAR

table_DAR = as.table(table_DAR)
write.csv(table_DAR, paste0(output_dir,"table_numbers_DA_Gut_KO_WT.csv"), row.names = TRUE, col.names = TRUE)

```

C. Plot data as a Volcano plot :

```{r}
# Add label of change
# 1. log2 Fold Change cut-off (create gene_change column):
DA_Gut_KO_WT$peak_change = "Unchanged"

DA_Gut_KO_WT$peak_change[Down_DAR] = "Down in KO"
DA_Gut_KO_WT$peak_change[Up_DAR] = "Up in KO"
```

```{r}
# 2.Plot Epidermis DAR data as Volcano plot ------------------
DA_Gut_KO_WT %>% 
  ggplot(aes(x=logFC, y=-log10(FDR))) +
  geom_point(alpha=0.75, size = 4, aes(colour=peak_change)) +
  scale_colour_manual(values = c("blue","grey","red")) +
  #geom_vline(xintercept = c(-0.5, 0.5), linetype="dashed", colour="grey") +
  geom_hline(yintercept = -(log10(0.05)), linetype="dashed", colour="grey") +
  ylim(0,25) + 
  xlim(-5.5,5.5) + 
  labs(x = "Log2 Fold Change") +
  labs(y = "-Log10(FDR)") +
  labs(colour = "Peak change") +
  ggtitle("ATAC Gut") +
  theme_bw(base_size = 18) +
  theme(panel.grid.minor = element_blank()) +
  theme(panel.grid.major = element_blank()) +
  theme(axis.text.x = element_text(angle = 0, colour = "black", size = 14)) +
  theme(axis.text.y = element_text(angle = 0, colour = "black", size = 14))
```

```{r}
# 3. Save Volcano plot: CHECK NAME CAREFULLY
outTitle <- paste0("ED_5b_ATAC_Gut_Volcano")
ggsave(paste0(output_dir, outTitle, ".pdf"),  width = 7, height = 5, units = "in", dpi = 72)
```

D. Export dataframe

```{r}
# 5. Exporting csv dataframe:
write.csv(DA_Gut_KO_WT, 
          file= paste0(output_dir,"Gut_DA_ATAC_KO_WT_labeled.csv"))
```

# Notes:

- Always check that output_dir is the correct destination before saving the file and check that export was successful.

```{r}
sessionInfo() 
```
